Limits on constraining galaxy profiles with current and future strong gravitational lensing

### 2013-02-15 limits/ms.tex
Wrote numerical methods section.

\documentclass[onecolumn,galley]{mn2e}
%\documentclass[onecolumn,galley,draft]{mn2e}
%\usepackage{myaasmacros}
\usepackage{amsmath}
\usepackage{graphicx}

\newcommand{\Glass}{{\sc Glass}}

\newcommand{\kpc}{\ensuremath{\mathrm{kpc}}}
\newcommand{\Msun}{\ensuremath{\mathrm{M}_\odot}}
\newcommand{\tabref}[1] {Table~\ref{#1}}
\newcommand{\figref}[1] {Figure~\ref{#1}}
\newcommand{\eqnref}[1] {Eq.~(\ref{#1})}
\newcommand{\secref}[1] {Section~\ref{#1}}
\newcommand{\Rmap}{\ensuremath{R_m}}
\newcommand{\Rpix}{\ensuremath{R_p}}
\newcommand{\e}[1]{\ensuremath{\times 10^{#1}}}
\renewcommand{\vec}[1]{\ensuremath{\boldsymbol{#1}}}

@@

The lensing equation
%
\begin{equation}
\vec\beta = \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta)
\label{lensing equation}
\end{equation}
%
maps an observed image position $\vec\theta$ to a source position $\vec\beta$
via the correction term 
%
\begin{equation}
\vec\alpha(\vec\theta) = \frac{4G}{c^2} \int \Sigma(\vec\theta')\frac{(\vec\theta - \vec\theta')}{\ |\vec\theta - \vec\theta'|^2}d^2\vec\theta'
\end{equation}
%
which accounts for the gravitational lensing effect of an intervening mass. For
our purposes we make use of the thin mass sheet approximation and assume that
the lensing mass is infinitely thin compared with the distance between the
source and the lens $D_{LS}$ and the lens and the observer $D_{L}$. These
distances are angular diameter distances where $D_* = (c/H_0)d_*$ and $d_*$ is
a dimensionless number that depends on cosmology.  The lens can then be thought
of as a projected surface density $\Sigma$ which diverts the path of a photon
instantaneously through the bending angle $\vec\alpha$.

Since we will later want to model the mass distribution with a computer it will
be convenient to choose units that make the relevant quantities of order unity.
We therefore measure lengths in light years, time in years, positions in
arcseconds, and choose $c=1$ and $4\pi G = N^2$, where $N^2 \equiv 206,265$
arcsec/rad. The mass unit is then $11.988\ \Msun$. It will also be useful to 
define a proxy to the Hubble constant $\nu \equiv N^2 H_0$.

The arrival (or travel) time $t$ of a photon depends not only on the geometric
path it takes, but also on the general relativistic gravitational time dilation
due to the lens at a redshift of $z_L$.  In its complete form, with all units
and cosmology included we can write the arrival time as
%
\begin{equation}
N^2ct(\vec\theta) = (1+z_L)\frac{D_{L}D_{S}}{D_{LS}}\frac12 |\vec\theta - \vec\beta|^2 - (1+z_L)\frac{4GD_{L}^2}{c^2}\int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| d^2\vec\theta'
\label{full arrival time}
\end{equation}
%
The non-standard factor of $D_L^2$ in the second term comes from the fact that $\Sigma$
has units of \Msun/lyr$^2$. We can clean this up by writing down a dimensionless time
delay 
%
\begin{equation}
\tau = \left[(1+z_L)D_{L}\right]^{-1}ctN^2 = \left[ (1+z_L) \frac{c}{H_0}d_L\right]^{-1}ctN^2 = \left[ (1+z_L) d_L\right]^{-1}\nu t
\label{tau}
\end{equation}
%
in terms of our previous definitions. If we
further define a dimensionless density
%
\begin{equation}
\kappa_\infty = \frac{4\pi G}{c^2}\frac{c}{H_0}d_L\Sigma
              = \frac{d_L}{\nu}\Sigma
\end{equation}
%
and a lensing potential
%
\begin{equation}
\psi(\vec\theta) = \frac1\pi \int \kappa_\infty(\vec\theta') \ln|\vec\theta - \vec\theta'| d^2\vec\theta'\
\end{equation}
%
we can express \eqnref{full arrival time} very compactly as
%
\begin{equation}
\tau(\vec\theta) = \frac12 \xi |\vec\theta-\vec\beta|^2 - \psi(\vec\theta)
\label{arrival time}
\end{equation}
%
where $\xi \equiv D_{S}/D_{LS}$. We explicitly write $\kappa_\infty$ to remind ourselves
that there is no source distance factor involved. This will be useful later when we consider
multiple sources.

If we think of $\tau$ as a surface then images will be observed where ever
$\nabla \tau = 0$. In other words, images obey Fermat's principle and appear at
extremums of the surface. From this we arrive back at \eqnref{lensing
equation}.

The only observables from \eqnref{arrival time} are the image positions, and
perhaps the stellar contribution to $\kappa_\infty$.  If the source object is
variable, however, then the observed light curves of each image will be
identical, but possibly shifted in time due to the different arrival times.
The time difference between identical portions of the light curves is called
the time delay $\Delta t$ between images.  Using \eqnref{tau} the time delay
$\Delta t_{21}$ between two images $\theta_1$ and $\theta_2$, where $\theta_2$
arrives after $\theta_1$, is $\Delta t_{21} = \Delta \tau_{21}(1+z_L)d_L /
\nu$. If $\beta$ and $\kappa_\infty$ were known exactly this would allow one
to precisely infer the Hubble constant. However, the source position is not
observable and the mass distribution is highly degenerate.

@@

Other degeneracies include ring degeneracies, where one is allowed to build
up a mass distribution as a series of mass rings; the presence of substructure;
and the effects of merging galaxies.

Shape degeneracy---a twisting in the mass.

@@

The lens configurations were generated within the new lens modeling framework
GLASS, described in Appendix \ref{GLASS Description}.

With this generated data, or data taken from real observations, GLASS can produce
an ensemble of possible mass distributions drawn from the space of solutions by
using a MCMC algorithm.  The technique is similar to that used in PixeLens
(cite Saha, Coles) but with an improved sampling algorithm.

As we have done in previous work, we imposed several priors on the reconstruction to
make it more physical, namely,
%
\begin{enumerate}
\item The density must be non-negative everywhere.
\item The density profile must have a slope $\le 0$. This prevents hollow rings.
\item The local gradient everywhere must point within $45^{\circ}$ of the center.
\item Density variations must be smooth.
\item Image parity.
\item The density is radially symmetric.
\end{enumerate}
%
Since lensing in this case is highly degenerate, many possible realizations are
produced. Each run produced an ensemble of 10000 models.

The quality of a model $M_i$ in an ensemble is judged by computing
\begin{equation}
\chi_i^2 = \frac{\sum_j (M_{ij} - G_j)^2}{\sum_j G_j^2}
\end{equation}
where $M$ and $G$ are the radial profiles of the model and galaxy, respectively.

@@

\section{Test Data}

@@

@@

>>>>

\documentclass[onecolumn,galley]{mn2e}
%\documentclass[onecolumn,galley,draft]{mn2e}
%\usepackage{myaasmacros}
\usepackage{mathrsfs}
\usepackage{amsmath}
\usepackage{graphicx}

\newcommand{\Glass}{{\sc Glass}}
\newcommand{\PixeLens}{{\sc PixeLens}}
\newcommand{\Rmap}{\ensuremath{R_\mathrm{map}}}
\newcommand{\Rpix}{\ensuremath{R_\mathrm{pix}}}
\newcommand{\M}{\ensuremath{\mathscr{M}}}
\newcommand{\E}{\ensuremath{\mathscr{E}}}

\newcommand{\kpc}{\ensuremath{\mathrm{kpc}}}
\newcommand{\Msun}{\ensuremath{\mathrm{M}_\odot}}
\newcommand{\tabref}[1] {Table~\ref{#1}}
\newcommand{\figref}[1] {Figure~\ref{#1}}
\newcommand{\eqnref}[1] {Eq.~(\ref{#1})}
\newcommand{\secref}[1] {Section~\ref{#1}}
\newcommand{\appref}[1] {Appendix~\ref{#1}}
\newcommand{\e}[1]{\ensuremath{\times 10^{#1}}}
\renewcommand{\vec}[1]{\ensuremath{\boldsymbol{#1}}}

@@

The lensing equation
%
\begin{equation}
\vec\beta = \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta)
\label{lensing equation}
\end{equation}
%
maps an observed image position $\vec\theta$ to a source position $\vec\beta$
via the correction factor 
%
\begin{equation}
\vec\alpha(\vec\theta) = \frac{4G}{c^2} \int \Sigma(\vec\theta')\frac{(\vec\theta - \vec\theta')}{\ |\vec\theta - \vec\theta'|^2}d^2\vec\theta'
\end{equation}
%
which accounts for the gravitational lensing effect of an intervening mass. For
our purposes we make use of the thin mass sheet approximation and assume that
the lensing mass is infinitely thin compared with the distance between the
source and the lens $D_{LS}$ and the lens and the observer $D_{L}$. These
distances are angular diameter distances where $D_* = (c/H_0)d_*$ and $d_*$ is
a dimensionless number that depends on cosmology.  The lens can then be thought
of as a projected surface density $\Sigma$ which diverts the path of a photon
instantaneously through the bending angle $\vec\alpha$.

Since we will later want to model the density distribution with a computer it will
be convenient to choose units that make the relevant quantities of order unity.
We therefore measure lengths in light years, time in years, positions in
arcseconds, and choose $c=1$ and $4\pi G = N^2$, where $N^2 \equiv 206,265$
arcsec/rad. The mass unit is then $11.988\ \Msun$. It will also be useful to 
define a proxy to the Hubble constant $\nu \equiv N^2 H_0$.

The arrival (or travel) time $t$ of a photon depends not only on the geometric
path it takes, but also on the general relativistic gravitational time dilation
due to the lens at a redshift of $z_L$.  In its complete form, with all units
and cosmology included we can write the arrival time as
%
\begin{equation}
N^2ct(\vec\theta) = (1+z_L)\frac{D_{L}D_{S}}{D_{LS}}\frac12 |\vec\theta - \vec\beta|^2 - (1+z_L)\frac{4GD_{L}^2}{c^2}\int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| d^2\vec\theta'
\label{full arrival time}
\end{equation}
%
The non-standard factor of $D_L^2$ in the second term comes from the fact that $\Sigma$
has units of \Msun/lyr$^2$. We can clean this up by writing down a dimensionless time
delay 
%
\begin{equation}
\tau = \left[(1+z_L)D_{L}\right]^{-1}ctN^2 = \left[ (1+z_L) \frac{c}{H_0}d_L\right]^{-1}ctN^2 = \left[ (1+z_L) d_L\right]^{-1}\nu t
\label{tau}
\end{equation}
%
in terms of our previous definitions. If we
further define a dimensionless density
%
\begin{equation}
\kappa_\infty = \frac{4\pi G}{c^2}\frac{c}{H_0}d_L\Sigma
              = \frac{d_L}{\nu}\Sigma
\end{equation}
%
and a lensing potential
%
\begin{equation}
\psi(\vec\theta) = \frac1\pi \int \kappa_\infty(\vec\theta') \ln|\vec\theta - \vec\theta'| d^2\vec\theta'\
\label{lensing potential}
\end{equation}
%
we can express \eqnref{full arrival time} very compactly as
%
\begin{equation}
\tau(\vec\theta) = \frac12 \xi |\vec\theta-\vec\beta|^2 - \psi(\vec\theta)
\label{arrival time}
\end{equation}
%
where $\xi \equiv D_{S}/D_{LS}$. We explicitly write $\kappa_\infty$ to remind ourselves
that there is no source distance factor involved. This will be useful later when we consider
multiple sources.

If we think of $\tau$ as a surface then images will be observed where ever
$\nabla \tau = 0$. In other words, images obey Fermat's principle and appear at
extremums of the surface. Note that $\nabla \tau$ is just \eqnref{lensing equation}.

The only observables from \eqnref{arrival time} are the image positions, and
perhaps the stellar contribution to $\kappa_\infty$.  If the source object is
variable, however, then the observed light curves of each image will be
identical, but possibly shifted in time due to the different arrival times.
The time difference between identical portions of the light curves is called
the time delay $\Delta t$ between images.  Using \eqnref{tau} the time delay
$\Delta t_{21}$ between two images $\vec\theta_1$ and $\vec\theta_2$, where $\vec\theta_2$
arrives after $\vec\theta_1$, is $\Delta t_{21} = \Delta \tau_{21}(1+z_L)d_L /
\nu$. If $\beta$ and $\kappa_\infty$ were known exactly this would allow one
to precisely infer the Hubble constant. However, the source position is not
observable and the density distribution is highly degenerate.

@@

Other degeneracies include ring degeneracies, where one is allowed to build
up a density distribution as a series of mass rings; the presence of substructure;
and the effects of merging galaxies.

Shape degeneracy---a twisting in the density.

@@

\subsection{A new lens modeling framework: \Glass}

\Glass\ is an evolution of the concepts employed by the free form modeling tool
\PixeLens\ but completely rewritten in Python and C. The key improvements are:
\begin{enumerate}
  \item A modular framework allows new priors to be added and modified easily;
  \item The basis functions approximating a model can be changed; 
  \item Non-linear constraints can be imposed in an automated post-processing step; 
  \item The central region can be have a higher resolution to capture steep models; 
  \item Stellar density can be used as an additional constraint.  
  \item Point or extended mass objects in the field.
\end{enumerate}
At its heart lies a new algorithm for sampling high dimensional linear spaces
that is used to generate lensing models. This algorithm was described and
tested in \cite{}. \Glass\ is also multi-threaded allowing it to run
efficiently on many-cored machines.  The software is freely available for
download. In our tests we will take advantage of each of these new features.

\subsection{Discrete models}

For this paper we will restrict ourselves to using the pixelated basis set that
has been employed in previous work and is implemented in \PixeLens. Here we will
briefly review the main ideas. The algorithm for generating models in \Glass\
samples a convex polytope in a high dimensional space whose interior points are
solutions to the lensing equation and satisfy other physically motivated linear
priors. We must therefore formulate all of our equations as equations linear in the
unknowns. We describe the density distribution
$\kappa$ as a set of discrete grid cells or pixels $\kappa_i$.
In this way, we rewrite the potential (\eqnref{lensing potential}) as
%
\begin{equation}
  \psi(\vec\theta) = \sum_n \kappa_n Q_n(\vec\theta)
\end{equation}
%
where the sum runs over all the pixels and $Q_n$ is the integral of the logarithm
over pixel $n$. The exact form for $Q$ is described in \appref{Q derivation}.
The pixels only cover a finite circular area with physical radius $\Rmap$ and 
pixel radius $\Rpix$. To account for any global shearing outside this region
from, e.g., a neighboring galaxy, we also add two shearing terms
%
\begin{equation}
\gamma_1(\theta_x^2 - \theta_y^2) + 2\gamma_2\theta_x\theta_y\quad.
\end{equation}
%
We can continue adding terms to account for assumed potentials or potentials
induced from the presence of other galaxy in the field. If the stellar density
has been estimated we can use this as a lower bound where the stellar potential
is a known constant
%
\begin{equation}
  \psi^*(\vec\theta) = \sum_n \kappa^*_n Q_n(\vec\theta)\quad.
\end{equation}
%
We can find the discretized lens equation by simply taking the gradient of the
above equations.

The solution space constrained by the lens equation and the arrival times are
not enough to form a closed volume. We have not imposed any strong constraints
on the density even so much as to ensure that it is non-negative everywhere. We
must therefore provide additonal linear constraints so that we generate physical
models. The explicit implementation for these priors has been explained in. In
summary, we require that
\begin{enumerate}
\item The density must be non-negative everywhere.
\item The density profile must have a slope everywhere $\le 0$. This prevents hollow rings.
\item The local gradient everywhere must point within $45^{\circ}$ of the center.
\item Density variations must be smooth.
\item Image parity is enforced.
\item The density is radially symmetric. (Optional)
\end{enumerate}

In the simplest form a single model for a lens is a tuple of the form
$\M = (\vec\kappa, \beta, \gamma_1, \gamma_2)$. A single model represents a 
single point in the solution space polytope. Using the MCMC sampling strategy
described in \cite{} we uniformly sample this space. Collectively, the sampled
models are referred to as an ensemble $\E = \{\M_i\}$, where we typically
generate 1000 models. We will typically show the ensemble average $|\E|$ in the
plots to follow.

@@

\section{Mock Galaxy Data}

@@

\subsection{Model quality}
We estimate of the quality of a model $\M_i$ by comparing the recovered
density profile $\rho_\M(r)$ against the profile of the mock galaxy $\rho_G(r)$.
In particular, the plots show the RMS quantity
%
\begin{equation}
  \chi_i = 100 \times \sqrt{\frac{\int_r (\rho_{\M_i}(r) - \rho_G(r))^2}{\int_r \rho_G(r)^2}}
  %\chi_i = \frac{\sum_r (\rho(r)_{\M_i} - \rho(r)_G)^2}{\sum_r \rho(r)_G^2}
\end{equation}

@@

\section{Derivation of pixelated density coefficients}
\label{Q derivation}
When the lens plane is pixelized we need a discrete form of the integral
%
\[\int \kappa(\vec\theta') \ln |\vec\theta-\vec\theta'| d^2\vec\theta' \]
%
In particular we want
%
\[\sum_n \kappa_n Q_n(\vec\theta - \vec\theta_n)\]
%
where $Q_n$ is the logarithm evaluated over the $n$th pixel at position $\vec\theta_n = (x_n, y_n)$. Let the pixel side length be $a$.
Instead of working with a position vector $\vec\theta$ we work in cartesian coordinates such that
%
$|\vec\theta| = r = \sqrt{x^2 + y^2}$. The integral now becomes
%
\[Q_n(x,y) = \frac12 \int_{y_-}^{y_+}\int_{x_-}^{x_+} \ln (x'^2+y'^2) dx' dy'\]
%
where $x_\pm = x_n \pm (a/2)$ and similarly for $y_\pm$.
Using the identity
%
\[\int \ln(x^2+y^2) dx = x \ln(x^2+y^2) - 2x + 2y\arctan(x/a) \]
%
we can express $Q_n$ as the sum of four parts
%
\[Q_n(x,y) = \frac12 \left[ \tilde Q_n(x_+,y_+)
        + \tilde Q_n(x_-,y_-)
        - \tilde Q_n(x_-,y_+)
        - \tilde Q_n(x_+,y_-) \right]\]
%
where
%
\[\tilde Q_n(x,y) = xy(\ln r^2 - 3) + x^2\arctan(y/x) + y^2\arctan(x/y)\]
